Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex

The splitting algorithm described in the previous subsections is
defined based on the local structure of \lang\ terms.  One artifact of
this design is that splitting generates

For example, the rule for splitting \bbonem\ case terms

% splitting rules are local term driven

% generate a lot of cruft

% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content

% recursive case would be recursive calls

% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff

% skip splitting.

% example is partition in 6.3

In order to be correct, any code at \bbonem\ must be split with the 
pessimistic assumption that it may result in work at the second stage.
When this assumption turns out not to be true---that is, purely monostage code---then 
splitting may produce second stage code which is needlessly costly 
(an example of this will occur in \ref{sec:exampleQS}).
Detecting and optimizing this case is in general a global program analysis,
since functions can be passed around as values.  
So instead of relying on hefty analysis, 
we take the approach of adding enough structure to the input language's type system to
allow the input code to {\em prove} itself to be---in some parts---monostage.

%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
%Yet even if a language designer desired a different implicit/explicit trade-off,
%then the three-world \lang\ would still be useful as a typed intermediate representation.

>>>>

The splitting algorithm described in the previous subsections operates
on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
argument into the boundary value, then decodes this tag in the
resumer. This logic occurs regardless of whether the terms forming the
branches of the \texttt{case} contain second stage computations.
Worse, if this \texttt{case} appeared in the body of a recursive
function with no other second stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:stagedsemantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.

The \texttt{partition} function in the \texttt{quickselect} example in
\ref{sec:exampleQS} is an example of such a function that contains no
second stage computations, but, if not defined within a $\pure$
annotation, would cause splitting to generate a recursive resumer.
